---
title: "GOV51 PSET3"
author: "\\small https://github.com/Steffny2004/gov51-ps3"
subtitle: "\\large Steffny Meegoda"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
	echo = TRUE,
	eval = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# 1 Setup and Data Exploration (10 points)

## 1.1 Load packages and data
Load the tidyverse, glmnet, and knitr packages. Then load the COMPAS dataset from data/compas_clean.csv.
```{r}
library(tidyverse)
library(glmnet)
library(knitr)
library(kableExtra)

compas <- read_csv("data/compas_clean.csv")
```

## 1.2 How many observations and variables are in the dataset? What is the unit of observation?
```{r}
n <- nrow(compas)
n_vars <- ncol(compas)
colnames(compas)
```

There are `n` observations and `n_vars` variables in the dataset, with the unit of observation being an individual who was convicted of a crime.

## 1.3 Examine the outcome variable
The variable recidivism equals 1 if the person was rearrested within two years and 0 otherwise.

(a) What is the overall recidivism rate in the data?
```{r}
mean(compas$recidivism)
```

The overall recidivism rate is 0.451, or 45.1%.

(b) Is this outcome continuous or binary? Why does this matter for how we interpret predictions from a linear model?
This outcome is binary, meaning that there are only 2 possible values for the outcome (0 or 1). As such, we interpret predictions from a linear model differently since continuous outcome interpretations can be described as "The outcome variable increases by $\beta_1 $ for a 1-unit increase in the predictor variable $x_i$, holding all other variables constant," where the continuous outcome could constitute any real value, whereas a binary represents an estimated rate/probability that falls strictly between 0 or 1 (inclusive). Because of this, a 1-unit increase in the predictor variable could result in our outcome variable being values that are not within its intended range of support.

## 1.4 Summary statistics
Create a summary statistics table showing the mean and standard deviation of all variables in the dataset. Briefly note which variables have the most variation and which have the least.
```{r}
summary_stats <- compas %>%

# Reshape table
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value") %>%
  
  # Calculate summary stats
  group_by(variable) %>%
  summarize(
    mean = round(mean(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    .groups = "drop")

# Create rendered table
summary_stats_rendered <- summary_stats %>%
  
  # Rename variables
  mutate(
    variable = recode(variable,
      recidivism = "Recidivism",
      age = "Age",
      female = "Female",
      black = "Black",
      hispanic = "Hispanic",
      asian = "Asian",
      other_race = "Other Race",
      native_american = "Native American",
      priors_count = "Priors Count",
      felony = "Felony",
      juv_fel_count = "Juvy Felony Count",
      juv_misd_count = "Juvy Misdemeanor Count",
      juv_other_count = "Juvy Other Count",
      has_juv_felony = "Has Juvy Felony",
      has_juv_misd = "Has Juvy Misdemeanor",
      age_under_25 = "Age Under 25", 
      age_25_45 = "Age Between 25-45"),
    
    
    # Order variables
    variable = factor(variable, levels = c(
      "Recidivism",
      "Age",
      "Female",
      "Black",
      "Hispanic",
      "Asian",
      "Other Race",
      "Native American",
      "Priors Count",
      "Felony",
      "Juvy Felony Count",
      "Juvy Misdemeanor Count",
      "Juvy Other Count",
      "Has Juvy Felony",
      "Has Juvy Misdemeanor",
      "Age Under 25", 
      "Age Between 25-45"))) %>%
  arrange(desc(sd)) %>%
  
  # Style table
  kable(
    booktabs = TRUE,
    caption = "Summary Statistics for COMPAS Recidivism Dataset",
    col.names = c("Variable", "Mean", "Std Dev"),
    align = c("l", "r", "r")) %>%
  kable_styling(latex_options = "hold_position")
      
# Display rendered table
summary_stats_rendered
  
```

From the above summary statistics table, it seems that `age` and `priors_count` have the most variation since their standard deviations are the largest, which makes sense since these are the only variables that can take on values greater than 1, as most of the other variables are binary indicators that only take on values of 0 or 1. The variables that have the smallest variation seem to be `asian` and `native_american`, which is likely because there are very few individuals in our dataset who fall into these categories compared to the other categories.

## 1.5 Explore key predictos
Create a bar chart showing the recidivism rate by number of prior offenses (`priors_count`). Group `priors_count` into categories: 0, 1–2, 3–5, 6–10, and 11+. What pattern do you see? 
```{r}
# Create categorical prior_counts var
compas_cat <- compas %>%
  mutate(priors_count_cat = 
           case_when(
             priors_count == 0 ~ "0", 
             priors_count >= 1 & priors_count <= 2 ~ "1-2", 
             priors_count >= 3 & priors_count <= 5 ~"3-5", 
             priors_count >= 6 & priors_count <= 10 ~ "6-10", 
             priors_count >= 11 ~ "11+")
         ) %>%
  group_by(priors_count_cat) %>%
  mutate(priors_count_cat = factor(priors_count_cat, 
                                   levels = c("0", "1-2", "3-5", "6-10", "11+"))) %>%
  summarize(recid_rate = mean(recidivism))

# Plot bar chart
ggplot(compas_cat, aes(x = priors_count_cat, y = recid_rate)) +
  geom_col(fill = "lightblue", color = "skyblue") +
  labs(
    title = "Recidivism Rate by Number of Prior Offenses",
    x = "Number of Prior Offenses",
    y = "Recidivism Rate"
  ) + 
  theme_minimal()
```

There is a clear pattern shown by the barplot that as the number of prior offenses increases, the recidivism rate also increases. Individuals with no prior offenses have the lowest recidivism rate, while those with many prior offenses have the highest rate, which suggests a strong positive relationship between a defendant's prior criminal history and the likelihood of recidivism

\newpage

# 2 OLS Baseline (15 points)

## 2.1 Train/test split
Set your seed to 51 for reproducibility. Split the data into a training set (80%) and a test set (20%).
```{r}
set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]
```

How many observations are in each set?
```{r}
n_train <- nrow(train)
n_test <- nrow(test)
```

There are `r n_train` observations in the training set and `n_test` observations in the test set.

## 2.2 Fit OLS with main effects only
Fit an OLS regression of recidivism on all other variables in the training data using `lm(recidivism ~ ., data = train)`. This gives you a model with 16 predictors.

(a) Which predictors are statistically significant at the 5% level?
```{r}
ols_simple <- lm(recidivism ~ ., data = train)
summary(ols_simple)
```

At the 5% level, it seems that the predictors `age`, `female`, `priors_count`, `felony`, `has_juv_misd`, and `age_25_45` are significant because their p-values are all less than 0.05.

(b) Interpret the coeﬀicient on priors_count in one sentence.
Holding all other variables constant, a 1-unit increase in `priors_count` is associated with a 0.0295 increase (2.95 percentage points) in recidivism rate.

## 2.3 In-sample vs. out-of-sample RMSE
Root Mean Squared Error (RMSE) measures average prediction error:
$$RMSE$$

(a) Compute the in-sample RMSE (predictions on the training data).
```{r}
# Compute train RMSE
pred_train_simple <- predict(ols_simple, train)
rmse_train_simple <- sqrt(mean((train$recidivism - pred_train_simple)^2))
rmse_train_simple
```

(b) Compute the out-of-sample RMSE (predictions on the test data).
```{r}
# Compute test RMSE
pred_test_simple <- predict(ols_simple, test)
rmse_test_simple <- sqrt(mean((test$recidivism - pred_test_simple)^2))
rmse_test_simple
```

(c) How large is the gap between in-sample and out-of-sample RMSE? Is OLS overfitting with 16 predictors?
```{r}
rmse_gap_simple <- rmse_test_simple - rmse_train_simple
rmse_gap_simple
```

The out-of-sample RMSE is larger than the in-sample RMSE by `r rmse_gap_simple`, which is very small and suggests that the OLS is a good fit with 16 predictors and is not overfit to the training data as it seems to generalize well on out-of-sample data.

## 2.4 Cross-validation by hand
A single train/test split depends on which observations happened to land in each group. Cross- validation averages over multiple splits to give a more stable estimate of out-of-sample performance.

Implement 5-fold cross-validation for the simple OLS model:

1. Randomly assign each training observation to one of 5 folds.
2. For each fold $k$ = 1, … , 5: fit OLS on the other 4 folds, predict fold $k$, compute RMSE.
3. Average the 5 RMSE values.

```{r}
# Assign folds
set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train)))

rmse_cv <- numeric(5)
for (k in 1:5) {
  train_k <- train[folds != k, ]
  test_k <- train[folds == k, ]
  model_k <- lm(recidivism ~
  ., data = train_k)
  pred_k <- predict(model_k, newdata = test_k)
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))
}

mean_rmse_cv <- mean(rmse_cv)
mean_rmse_cv
```

(a) What is the 5-fold CV estimate of RMSE? How does it compare to your single-split out-of-sample RMSE?
The 5-fold cross-validation estimate of RMSE is `mean_rmse_cv` = 0.4620. This is extremely close to the single-split out-of-sample RMSE of `rmse_test_simple` = 0.4622, with a difference of about 0.0002, which suggests that the model’s out-of-sample performance is similar and does not depend much on one particular train/test split.

(b) In your own words, why is cross-validation more reliable than a single train/test split?
Cross-validation is more reliable because it trains the model on all of the data points, since each fold of data will be used for training 4 times and used for testing 1 time, and then averages across all splits This is better than just a single train/test split because depending on how the initial data is split, the single-split test data sample may not be representative of the overall data. So, your test RMSE could appear much higher than your training RMSE and suggest overfitting when that's not the reality because you tested your model on an odd portion of your data. Averaging over multiple splits can reduce some of that randomness/variation 

(c) When we use `cv.glmnet()` later, it does this same process internally to choose the penalty parameter $\lambda$. Why is cross-validation the right tool for choosing $\lambda$?
Cross-validation is the right tool for choosing $\lambda$ because it evaluates model performance on the unseen test data, allowing us to select the value of $\lambda$ that minimizes out-of-sample prediction error. Since $\lambda$ controls the complexity of the model, cross-validation helps balance both overfitting and underfitting, which leads to a model that generalizes well to new data.


\newpage

# 3 The Kitchen Sink: What Happens When You Add Too Many Predictors? (15 points)
The simple OLS model uses 16 main effects. But maybe interactions between variables would help. What if being young AND having prior offenses is different from the sum of each effect alone? Let’s find out what happens when we throw in *everything.*

## 3.1 Build the expanded model
The formula below creates all pairwise interactions plus squared terms for the continuous variables. This expands our feature set from 16 to about 141 predictors.
```{r}
# Full interaction formula
f_full <- recidivism ~ (.)^2 +
  I(age^2) + I(priors_count^2) +
  I(juv_fel_count^2) + I(juv_misd_count^2) +
  I(juv_other_count^2)

# Create design matrices for glmnet (we'll need these later)
X_train <- model.matrix(f_full, data = train)[, -1]
X_test <- model.matrix(f_full, data = test)[, -1]
y_train <- train$recidivism
y_test <- test$recidivism

# How many features now?
ncol(X_train)
```

(a) How many predictors does this expanded model have?
The expanded model now has 141 predictors

(b) What is the ratio of training observations to predictors? (Compare to the simple model.)
```{r}
# Ratio for expanded model
ratio_expanded <- nrow(X_train) / ncol(X_train)

# Ratio for simple model
ratio_simple <- nrow(train) / (ncol(train) - 1)

# Display results
cat("Ratio for Expanded Model:", ratio_expanded, 
    "\nRatio for Simple Model:", ratio_simple)
```

From above, we see that the ratio for the expanded model of 40.93 is much smaller than the simpler model ratio of 360.69. This means that relative to the number of predictors, the expanded model has far less data per variable, which makes it more likely to overfit.

## 3.2 Fit OLS with the kitchen sink
Fit OLS on the expanded feature set: ols_full <- lm(f_full, data = train).
```{r}
ols_full <- lm(f_full, data = train)
```

(a) Compute the in-sample RMSE for this model.
```{r}
# Compute train RMSE
pred_train_full <- predict(ols_full, train)
rmse_train_full <- sqrt(mean((train$recidivism - pred_train_full)^2))
rmse_train_full

```


(b) Compute the out-of-sample RMSE.
```{r}
# Compute train RMSE
pred_test_full <- predict(ols_full, test)
rmse_test_full <- sqrt(mean((test$recidivism - pred_test_full)^2))
rmse_test_full
```

(c) How does the in-sample/out-of-sample gap compare to the simple 16-predictor OLS? What happened and why?
```{r}
rmse_gap_full <- rmse_test_full - rmse_train_full
rmse_gap_full
```

The gap between in-sample and out-of-sample RMSE is 0.04, which is much larger for the expanded model compared to the simple model that had a gap of 0.006. This could suggest that the expanded model overfits the training data because by including many interaction and nonlinear terms, the model becomes too flexible and "wiggly" to capture noise in the training data, leading to worse performance on the unseen, new test data.

## 3.3 The overfitting lesson
In 2–3 sentences, explain why adding 180 interaction terms made the model’s **in-sample** fit better but its **out-of-sample** predictions worse. What is this phenomenon called?
Adding 180 interaction terms made the model be extremely well-tuned to the training data and captures the data better, but this means that it is so well-tuned to that data that it won't do as well predicting on new data. Also, the 180 interaction terms is overkill, since it's likely that not all these terms are significant either. This phenomenon is called overfitting because the model overfit to the noise of the training data by using too many predictors and lost the signal, so it performed poorly on the test data that it's never seen before and couldn't identify the general pattern of this new data.

\newpage 

# 4 Ridge Regression: Shrink Everything (15 points)

Ridge regression adds an L2 penalty that shrinks all coeﬀicients toward zero. Larger $\lambda$ means more shrinkage. The key idea: Ridge trades a small amount of bias for a large reduction in variance.

## 4.1 Fit Ridge with cross-validation
Use `cv.glmnet()` with `alpha = 0` to fit Ridge regression on the expanded training data (the X_train matrix with ~141 features). Use 10-fold cross-validation (the default) to select the optimal $\lambda$.
```{r}
set.seed(51) # for reproducible CV folds
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
```

(a) What is the optimal$\lambda$ selected by cross-validation? (Use `lambda.min`)
```{r}
# Find optimal lambda
lambda_ridge <- cv_ridge$lambda.min

# Display results
cat("The optimal lambda is", lambda_ridge)
```

(b) Plot the cross-validation curve using `plot(cv_ridge)`. What does this plot show?
```{r}
plot(cv_ridge)
```

This plot shows the cross-validated mean squared error (MSE) as a function of $-\log(\lambda)$. Each point represents the average validation error across folds, and the vertical dashed lines indicate the range of $\lambda$ that minimizes the error (`lambda.min` falls within these dashed lines since ). The numbers at the top represent the number of nonzero coefficients, which remains constant at 128 for Ridge This indicates that even though the total number of variables in our Ridge model should be 141 (since Ridge doesn't set any coefficients to 0), there are likely interaction terms that are always 0 by default, such as `black:hispanic`, `asian:other_race`, etc., since these predictors are mutually exclusive.

## 4.2 Ridge RMSE
Compute the out-of-sample RMSE for Ridge. Compare to both the simple OLS (16 vars) and the kitchen-sink OLS (~141 vars). What happened?
```{r}
# Compute test RMSE
pred_test_ridge <- predict(cv_ridge, s = lambda_ridge, newx = X_test)
rmse_test_ridge <- sqrt(mean((y_test - pred_test_ridge)^2))

# Display results
cat("Test RMSE:", rmse_test_ridge)
```

The out-of sample RMSE for Ridge is 0.470, which is slightly lower than the out-of sample RMSE's for the kitchen-sink model (`r rmse_test_full`), but higher than the simple model (`r rmse_test_simple`). This suggests that Ridge performs better than the kitchen-sink mode by reducing overfitting through shrinking coefficients, but while Ridge's regularization helps control the complexity of the expanded model, the simpler model still provides the best out-of-sample performance in this case.

## 4.3 Did Ridge set any coeﬀicients to zero?
Extract the Ridge coeﬀicients using `coef(cv_ridge, s = "lambda.min").` Did Ridge eliminate any variables? Why or why not?
```{r}
# Count number of retained variables
ridge_coefs <- coef(cv_ridge, s = "lambda.min")
num_nonzero_ridge <- sum(ridge_coefs[-1] != 0)

# Display results
cat("Number of Retained Vars:", num_nonzero_ridge)
```

There are 128 nonzero coefficients kept by Ridge, so 13 were "eliminated". Even though Ridge does not set any coefficients exactly to 0, some coefficients appear as zero or missing are likely due to redundancy or multicollinearity in the data, rather than Ridge performing variable selection. As such, these predictors are considered less important or not as highly associated with the outcome than other predictors. Additionally, as I discussed earlier, there are likely interaction terms that are always 0 by default, such as `black:hispanic`, `asian:other_race`, etc., since these predictors are mutually exclusive, which causes these to have a coefficient of 0.

\newpage

# 5 LASSO Regression: Shrink and Select (15 points)
LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty. Unlike Ridge, LASSO can shrink coeﬀicients all the way to exactly zero, effectively performing variable selection.

## 5.1 Fit LASSO with cross-validation
Use `cv.glmnet()` with alpha = 1 on the expanded training data. Remember to set your seed first for reproducibility.

```{r}
set.seed(51)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
```

(a) What is the optimal $\lambda$?
```{r}
# Find optimal lambda
lambda_lasso <- cv_lasso$lambda.min

# Display results
cat("The optimal lambda is", lambda_lasso)
```

(b) Plot the cross-validation curve.
```{r}
plot(cv_lasso)
```
Again, this plot shows the cross-validated mean squared error (MSE) as a function of $-\log(\lambda)$ but now for our LASSO model. Each point represents the average validation error across folds, and the vertical dashed lines indicate the range of $\lambda$ that minimizes the error (`lambda.min` falls within these dashed lines). The numbers at the top represent the number of nonzero coefficients, which shows that as $-\log(\lambda)$ increases (meaning that $\lambda$ decreases along the x-axis), then the number of nonzero coefficients increases, which makes sense conceptually since $\lambda$ decreasing means a less harsh penalty and more coefficients are allowed into the model. In our optimal $\lambda$ range, we see that the model retains about ~27 coefficients.

## 5.2 LASSO variable selection
Extract the LASSO coeﬀicients at lambda.min.
(a) How many of the ~141 variables survived (non-zero coeﬀicient)? How many were eliminated?
```{r}
# Count number of retained variables
lasso_coefs <- coef(cv_lasso, s = "lambda.min")
num_nonzero_lasso <- sum(lasso_coefs[-1] != 0)

# Display results
cat("Number of Retained Vars:", num_nonzero_lasso)
```
LASSO selected 27 predictors and eliminated the rest of the 114 predictors. After examining the coefficient matrix, there seemed to be a tendency to keep predictors and interactions between `female`, `age`, `age_under_25`, `priors_count`, `felony`, and categories that tended to relate to juvy.  

(b) List a few of the surviving interaction terms. Do they make substantive sense?
A few of the surviving interaction terms are `juv_fel_count:age_under_25`, `juv_other_count:age_under_25`,  `has_juv_felony:age_under_25`, which logistically makes sense since these terms are saying that for individuals under 25, the effect for having a juvy felony or their juvy felony count increasing is greater than if they were in the 25-45 age category. Other categories show interactions between racial groups and having certain charges like `hispanic:juv_fel_count` and `black:juv_fel_count`, both of which have a positive effect and suggest differences between racial groups in the way charges are distributed

(c) Baker (2025) argues that LASSO reduces “subjective discretion” in model building. How does what you see here illustrate that argument?
LASSO appears to reduce “subjective discretion” by only keeping predictors based on their contribution to out-of-sample predictive performance to minimize the CV error, whereas manually choosing which predictors to keep might reflect some bias on the researcher's part by either not including predictors that might actually have an effect (omitted variable bias), or, including variables that are not that significant but improve other metrics of the model suhc as $R^2$. As such, LASSO algorithmically and optimally chooses which predictors to keep rather than a researcher using solely their theoretical framework to decide.

## 5.3 LASSO RMSE
Compute the out-of-sample RMSE for LASSO. How does it compare to OLS (both versions) and Ridge?
```{r}
# Compute test RMSE
pred_test_lasso <- predict(cv_lasso, s = lambda_lasso, newx = X_test)
rmse_test_lasso <- sqrt(mean((y_test - pred_test_lasso)^2))

# Display results
cat("Test RMSE:",  rmse_test_lasso)
```

The out-of sample RMSE for LASSO is 0.459, making it the best performing model with the smallest out-of sample RMSE value in comparison to the simple model (0.462), the kitchen-sink model (0.489), and even the Ridge model (0.470).

\newpage

# 6 Model Comparison and Interpretation (30 points)

## 6.1 Elastic Net
Fit an Elastic Net with `alpha = 0.5` on the expanded training data. Remember to set your seed first. Compute its out-of-sample RMSE and count how many variables it retains.

```{r}
# Fit Elastic Net
set.seed(51)
cv_enet <- cv.glmnet(X_train, y_train, alpha = 0.5, nfolds = 10)

# Find optimal lambda
lambda_enet <- cv_enet$lambda.min

# Compute test RMSE
pred_test_enet <- predict(cv_enet, s = lambda_enet, newx = X_test)
rmse_test_enet <- sqrt(mean((y_test - pred_test_enet)^2))

# Count number of retained variables
enet_coefs <- coef(cv_enet, s = "lambda.min")
num_nonzero_enet <- sum(enet_coefs[-1] != 0)

# Display results
cat("Test RMSE:", rmse_test_enet, "\nNumber of Retained Vars:", num_nonzero_enet)
```

## 6.2 The big comparison table
Fill in this table with your results:
```{r}
comparison_table <- tibble(
  Model = c(
    "OLS (main effects. 16 vars)",
    "OLS (kitchen sink, ~141 vars)",
    "Ridge",
    "LASSO",
    "Elastic Net"
  ),
  Predictors = c(
    16,
    ncol(X_train),
    "All (~128)",
    num_nonzero_lasso,
    num_nonzero_enet
  ),
  RMSE = c(
    rmse_test_simple,
    rmse_test_full,
    rmse_test_ridge,
    rmse_test_lasso,
    rmse_test_enet
  )
)

comparison_table %>%
  mutate(RMSE = round(RMSE, 3)) %>%
  kable(
    col.names = c("Model", "Predictors", "Out-of-Sample RMSE"),
    align = c("l", "c", "c"),
    booktabs = TRUE
  ) %>%
  kable_styling(latex_options = "hold_position")

```

(a) Which model has the lowest out-of-sample RMSE?
The Elastic Net model has the lowest out-of-sample RMSE of 0.458, making it the best-performing model in terms of predictive accuracy.

(b) Which model has the best in-sample fit? Is that the same model?
The kitchen-sink OLS model has the best in-sample fit because it has the lowest training RMSE of 0.4890411. However, it does not perform best out-of-sample since it overfits the training data.

(c) In 2–3 sentences, explain what this table teaches about the bias-variance tradeoff.
This table shows the bias-variance tradeoff through the differences in out-of-sample RMSE performance based on model. The kitchen-sink OLS model has low bias but high variance due to its high model complexity, leading to overfitting and poor out-of-sample performance. However, regularized models like Ridge, LASSO, and Elastic Net introduce some bias by shrinking coefficients, but overall reduce variance and improve generalization, which results in the better out-of-sample performance that we see in our table.

## 6.3 Baker article connection
In Baker (2025), two experts in the *Halliburton* securities case selected different peer firms and reached different conclusions about whether a stock drop was statistically significant.

(a) How did LASSO help narrow the disagreement between dueling experts?
LASSO helped narrow the disagreement by using an algorithmic and more computational method of selecting peer firms rather than relying on the subjective choices made by each expert. This means that instead of manually choosing which firms to include, which can lead to different conclusions, LASSO automatically selects the firms that best predict the outcome based on cross-validation. This reduces the scope for disagreement because both experts can rely on the same objective selection procedure.

(b) Based on your experience with LASSO in this problem set, what aspect of the traditional OLS approach gives the analyst the most room for manipulation?
The traditional OLS approach allows the most room for manipulation because of the subjective selection of variables or peer firms included in the model. Analysts can choose which predictors to include or exclude, and these choices can significantly affect the results and lead to different conclusions. LASSO, however, reduces this subjective discretion by using a systematic approach to variable selection.

## 6.4 Prediction vs. explanation
We used these models to predict `recidivism.` Suppose instead you wanted to know whether `priors_count` causes higher recidivism.

(a) Would the LASSO coeﬀicient on `priors_count` answer this causal question? Why or why not?
No, the LASSO coefficient on `priors_count` would not answer the causal question. This is because LASSO is better for prediction, not causal inference, as its coefficient estimates are biased due to regularization. Additionally, the model does not account for potential confounding variables or omitted variable bias, so the estimated relationship can only reflect correlation an not causation.

(b) What is the difference between using regression for prediction versus using it for causal inference?
The difference between using regression for prediction versus using it for causal inference is that prediction focuses on accurately forecasting outcomes, while causal inference aims to estimate the effect of one variable on another. So, predictive models like LASSO prioritize minimizing the prediction error and can include biased coefficients or omit variables that are important for interpretation, while causal inference depends on model specification to determine the true causal effect, ensuring that estimates are not driven by confounding factors.

## 6.5 Algorithmic fairness
Algorithms like COMPAS are used in real courtrooms to inform bail and sentencing decisions.

(a) Using your simple OLS model (16 vars), compute the mean predicted recidivism probability separately for Black defendants and white defendants in the test set. Is there a difference?
```{r}
# Predict
pred_test_simple <- predict(ols_simple, test)
test_with_preds <- test %>%
  mutate(pred = pred_test_simple)

# Define groups
black_group <- test_with_preds %>% 
  filter(black == 1)

white_group <- test_with_preds %>%
  filter(black == 0 & hispanic == 0 & asian == 0 & other_race == 0 & native_american == 0)

# Compute mean predicted recidivism within groups
mean_pred_black <- mean(black_group$pred)
mean_pred_white <- mean(white_group$pred)

# Display results
cat("Mean Predicted Recidivism Probability:",
    "\nBlack Defendants:", mean_pred_black, 
    "\nWhite Defendants:", mean_pred_white)
```

The mean predicted recidivism probability for Black defendants is 0.512, while for white defendants it is 0.379. There is a substantial difference between the two groups, with Black defendants having a higher predicted probability of recidivism by about 13 percentage points.
