---
title: "GOV51 PSET3"
author: "\\small https://github.com/Steffny2004/gov51-ps3"
subtitle: "\\large Steffny Meegoda"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
	echo = TRUE,
	eval = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# 1 Setup and Data Exploration (10 points)

## 1.1 Load packages and data
Load the tidyverse, glmnet, and knitr packages. Then load the COMPAS dataset from data/compas_clean.csv.
```{r}
library(tidyverse)
library(glmnet)
library(knitr)
library(kableExtra)

compas <- read_csv("data/compas_clean.csv")
```

## 1.2 How many observations and variables are in the dataset? What is the unit of observation?
```{r}
n <- nrow(compas)
n_vars <- ncol(compas)
colnames(compas)
```

There are `n` observations and `n_vars` variables in the dataset, with the unit of observation being an individual who was convicted of a crime.

## 1.3 Examine the outcome variable
The variable recidivism equals 1 if the person was rearrested within two years and 0 otherwise.

(a) What is the overall recidivism rate in the data?
```{r}
rec_rate <- mean(compas$recidivism)
rec_rate
```

The overall recidivism rate is `rec_rate`.

(b) Is this outcome continuous or binary? Why does this matter for how we interpret predictions from a linear model?
This outcome is binary, meaning that there are only 2 possible values for the outcome (0 or 1). As such, we interpret predictions from a linear model differently since continuous outcome interpretations can be described as "The outcome variable increases by $\beta_1 $ for a 1-unit increase in the predictor variable $x_i$, holding all other variables constant," where the continuous outcome could constitute any real value, whereas a binary represents an estimated rate/probability that falls strictly between 0 or 1 (inclusive). Because of this, a 1-unit increase in the predictor variable could result in our outcome variable being values that are not within its intended range of support.

## 1.4 Summary statistics
Create a summary statistics table showing the mean and standard deviation of all variables in the dataset. Briefly note which variables have the most variation and which have the least.
```{r}
summary_stats <- compas %>%

# Reshape table
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value") %>%
  
  # Calculate summary stats
  group_by(variable) %>%
  summarize(
    mean = round(mean(value, na.rm = TRUE), 2),
    sd = round(sd(value, na.rm = TRUE), 2),
    .groups = "drop")

# Create rendered table
summary_stats_rendered <- summary_stats %>%
  
  # Rename variables
  mutate(
    variable = recode(variable,
      recidivism = "Recidivism",
      age = "Age",
      female = "Female",
      black = "Black",
      hispanic = "Hispanic",
      asian = "Asian",
      other_race = "Other Race",
      native_american = "Native American",
      priors_count = "Priors Count",
      felony = "Felony",
      juv_fel_count = "Juvy Felony Count",
      juv_misd_count = "Juvy Misdemeanor Count",
      juv_other_count = "Juvy Other Count",
      has_juv_felony = "Has Juvy Felony",
      has_juv_misd = "Has Juvy Misdemeanor",
      age_under_25 = "Age Under 25", 
      age_25_45 = "Age Between 25-45"),
    
    
    # Order variables
    variable = factor(variable, levels = c(
      "Recidivism",
      "Age",
      "Female",
      "Black",
      "Hispanic",
      "Asian",
      "Other Race",
      "Native American",
      "Priors Count",
      "Felony",
      "Juvy Felony Count",
      "Juvy Misdemeanor Count",
      "Juvy Other Count",
      "Has Juvy Felony",
      "Has Juvy Misdemeanor",
      "Age Under 25", 
      "Age Between 25-45"))) %>%
  arrange(variable) %>%
  
  # Style table
  kable(
    booktabs = TRUE,
    caption = "Summary Statistics for COMPAS Recidivism Dataset",
    col.names = c("Variable", "Mean", "Std Dev"),
    align = c("l", "r", "r")) %>%
  kable_styling(latex_options = "hold_position")
      
# Display rendered table
summary_stats_rendered
  
```

From the above summary statistics table, it seems that `age` and `priors_count` have the most variation since their standard deviations are the largest, which makes sense since these are the only continuous variables in the dataset that can take on values greater than 1, as most of the other variables are binary indicators that only take on values of 0 or 1. The variables that have the smallest variation seem to be `asian` and `native_american`, which is likely because there are very few individuals in our dataset who fall into these categories comapred to the other categories.

## 1.5 Explore key predictos
Create a bar chart showing the recidivism rate by number of prior offenses (`priors_count`). Group `priors_count` into categories: 0, 1–2, 3–5, 6–10, and 11+. What pattern do you see? 
```{r}
# Create categorical prior_counts var
compas_cat <- compas %>%
  mutate(priors_count_cat = 
           case_when(
             priors_count == 0 ~ "0", 
             priors_count >= 1 & priors_count <= 2 ~ "1-2", 
             priors_count >= 3 & priors_count <= 5 ~"3-5", 
             priors_count >= 6 & priors_count <= 10 ~ "6-10", 
             priors_count >= 11 ~ "11+")
         )

# Plot bar chart
ggplot(compas_cat, mapping = aes(x = priors_count_cat, y = recidivism / length(priors_count_cat))) +
  geom_col(position = "fill") + 
  labs(
    title = "Recidivism Rate by Number of Prior Offenses", 
    x = "Number of Prior Offenses", 
    y = "Recidivism Rate"
    ) + 
  theme_minimal()
```

\newpage

# 2 OLS Baseline (15 points)

## 2.1 Train/test split
Set your seed to 51 for reproducibility. Split the data into a training set (80%) and a test set (20%).
```{r}
set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]
```

How many observations are in each set?
```{r}
n_train <- nrow(train)
n_test <- nrow(test)
```

There are `n_train` observations in the training set and `n_test` observations in the test set.

## 2.2 Fit OLS with main effects only
Fit an OLS regression of recidivism on all other variables in the training data using `lm(recidivism ~ ., data = train)`. This gives you a model with 16 predictors.

(a) Which predictors are statistically significant at the 5% level?
```{r}
ols_simple <- lm(recidivism ~ ., data = train)
summary(ols_simple)
```

At the 5% level, it seems that the predictors `age`, `female`, `priors_count`, `felony`, and `age_25_45` are significant.

(b) Interpret the coeﬀicient on priors_count in one sentence.
On average, for an individual that is 0 years old, white, male, was not in juvy, and did not commit a felony, as the number of prior offenses that they have increases by 1, then their recidivism rate increase by 0.015761 (1.5761%).

## 2.3 In-sample vs. out-of-sample RMSE
Root Mean Squared Error (RMSE) measures average prediction error:
$$RMSE$$

(a) Compute the in-sample RMSE (predictions on the training data).
```{r}
# Compute train RMSE
pred_train_simple <- predict(ols_simple, train)
rmse_train_simple <- sqrt(mean((compas$recidivism - pred_train_simple)^2) / n_train)
rmse_train_simple
```

(b) Compute the out-of-sample RMSE (predictions on the test data).
```{r}
# Compute test RMSE
pred_test_simple <- predict(ols_simple, test)
rmse_test_simple <- sqrt(mean((compas$recidivism - pred_test)^2) / n_test)
rmse_test_simple
```

(c) How large is the gap between in-sample and out-of-sample RMSE? Is OLS overfitting with 16 predictors?
```{r}
rmse_gap_simple <- rmse_test_simple - rmse_train_simple
rmse_gap_simple
```

The out-of-sample RMSE is larger than the in-sample RMSE by `rmse_test_simple`, which is very small and suggests that the OLS is a good fit with 16 predictors and is not actually overfit to the training data.

## 2.4 Cross-validation by hand
A single train/test split depends on which observations happened to land in each group. Cross- validation averages over multiple splits to give a more stable estimate of out-of-sample performance.

Implement 5-fold cross-validation for the simple OLS model:

1. Randomly assign each training observation to one of 5 folds.
2. For each fold $k$ = 1, … , 5: fit OLS on the other 4 folds, predict fold $k$, compute RMSE.
3. Average the 5 RMSE values.

```{r}
# Assign folds
set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train)))

rmse_cv <- numeric(5)
for (k in 1:5) {
  train_k <- train[folds != k, ]
  test_k <- train[folds == k, ]
  model_k <- lm(recidivism ~
  ., data = train_k)
  pred_k <- predict(model_k, newdata = test_k)
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))
}

mean(rmse_cv)
```

(a) What is the 5-fold CV estimate of RMSE? How does it compare to your single-split out-of-sample RMSE?
The 5-fold CV estimate of RMSE is 0.4584038, which is smaller than our single-split out-of-sample RMSE of `test_rmse`. 

(b) In your own words, why is cross-validation more reliable than a single train/test split?
Cross-validation is more reliable because it trains the model on all of the data points eventually, since essentially each fold of data will be used for training 4 times and used for testing 1 time. This is better than just a single train/test split because depending on how the initial data is split, the test data sampple may not be entirely representative of the overall data, so your test RMSE could appear much higher than your training RMSE and suggest overfitting when that's not the reality and you actually tested your model on an odd portion of your data.

(c) When we use `cv.glmnet()` later, it does this same process internally to choose the penalty parameter $\lambda$. Why is cross-validation the right tool for choosing $\lambda$?
Cross-validation is the right tool for choosing $\lambda$ because just as overiftting helps us tune and optimize the model's coefficients such that our training RMSE and test RMSE are low, it also helps us tune and optimize the $\lambda$ paramter in our model to minimize the RMSE's instead of just picking 1 value for $\lambda$ and sticking with it when it might not be the best choice for our final model.

\newpage

# 3 The Kitchen Sink: What Happens When You Add Too Many Predictors? (15 points)
The simple OLS model uses 16 main effects. But maybe interactions between variables would help. What if being young AND having prior offenses is different from the sum of each effect alone? Let’s find out what happens when we throw in *everything.*

## 3.1 Build the expanded model
The formula below creates all pairwise interactions plus squared terms for the continuous variables. This expands our feature set from 16 to about 141 predictors.
```{r}
# Full interaction formula
f_full <- recidivism ~ (.)^2 +
  I(age^2) + I(priors_count^2) +
  I(juv_fel_count^2) + I(juv_misd_count^2) +
  I(juv_other_count^2)

# Create design matrices for glmnet (we'll need these later)
X_train <- model.matrix(f_full, data = train)[, -1]
X_test <- model.matrix(f_full, data = test)[, -1]
y_train <- train$recidivism
y_test <- test$recidivism

# How many features now?
ncol(X_train)
```

(a) How many predictors does this expanded model have?
The expanded model now has 209 features.

(b) What is the ratio of training observations to predictors? (Compare to the simple model.)
```{r}
# Ratio for expanded model
ratio_expanded <- nrow(X_train) / ncol(X_train)

# Ratio for simple model
ratio_simple <- nrow(train) / (ncol(train) - 1)

# Display results
cat("Ratio for Expanded Model:", ratio_expanded, 
    "\nRatio for Expanded Model:", ratio_simple)
```

From above, we see that the ratio for the expanded model is much larger at `ratio_expanded` compared to the simpler model at `ratio_simple`.

## 3.2 Fit OLS with the kitchen sink
Fit OLS on the expanded feature set: ols_full <- lm(f_full, data = train).
```{r}
ols_full <- lm(f_full, data = train)
```

(a) Compute the in-sample RMSE for this model.
```{r}
# Compute train RMSE
pred_train_full <- predict(ols_full, as.data.frame(X_train))
rmse_train_full <- sqrt(mean((y_train - pred_train_full)^2) / nrow(X_train))
rmse_train_full
```


(b) Compute the out-of-sample RMSE.
```{r}
# Compute train RMSE
pred_test_full <- predict(ols_full, as.data.frame(X_test))
rmse_test_full <- sqrt(mean((y_test - pred_train_full)^2) / nrow(X_test))
rmse_test_full
```

(c) How does the in-sample/out-of-sample gap compare to the simple 16-predictor OLS? What happened and why?
```{r}
rmse_gap_full <- rmse_test_full - rmse_train_full
rmse_gap_full
```

The out-of-sample RMSE is larger than the in-sample RMSE by `rmse_gap_full`, which is again very small and comparable to the simple 19-predictor OLS of `rmse_test_simple`, which suggests that the OLS is a good fit with 16 predictors and is not actually overfit to the training data.

## 3.3 The overfitting lesson
In 2–3 sentences, explain why adding 180 interaction terms made the model’s **in-sample** fit better but its **out-of-sample** predictions worse. What is this phenomenon called?
Adding 180 interaction terms made the model be extremely well-tuned to the training data and captures the data almost perfectly, but this means that it is so well-tuned to that data that it won't do as well predicting on new data. Also, the 180 interaction terms is overkill, since it's likely that not all these terms are significant either. This phenomenon is called overfitting because the model overfit to the noise of the training data and lost the signal, so it performed poorly on the test data that it's never seen before and couldn't identify the general pattern of this new data.

\newpage 

# 4 Ridge Regression: Shrink Everything (15 points)

Ridge regression adds an L2 penalty that shrinks all coeﬀicients toward zero. Larger $\lambda$ means more shrinkage. The key idea: Ridge trades a small amount of bias for a large reduction in variance.

## 4.1 Fit Ridge with cross-validation
Use `cv.glmnet()` with `alpha = 0` to fit Ridge regression on the expanded training data (the X_train matrix with ~141 features). Use 10-fold cross-validation (the default) to select the optimal $\lambda$.
```{r}
set.seed(51) # for reproducible CV folds
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
```

(a) What is the optimal$\lambda$ selected by cross-validation? (Use `lambda.min`)
```{r}
lambda_ridge <- cv_ridge$lambda.min
lambda_ridge
```
The optimal $\lambda$ is `lambda_ridge`

(b) Plot the cross-validation curve using `plot(cv_ridge)`. What does this plot show?
```{r}
plot(cv_ridge)
```

This plot is showing the Mean-Squared Error values that are calculated from the Ridge model for different values of $\lambda$ on the $-log(\lambda)$ scale. The 195 at the top of the plot is the number of non-zero coefficients as the value of $-log(\lambda)$ increases, and it makes sense that this number is consistently 195 since Ridge only makes it so that some coefficients are approaching 0 rather than setting them to exactly 0. The two vertical dashed lines on the groph indicate the region of $-log(\lambda)$ values that provide the best model performance.  From this plot, it appears that the optimal $-log(\lambda)$ is between 0 and 1.5 (meaning that $\lambda \approx e^{-1} \approx 0.3679$, which is close to the optimal $\lambda$ value that we found in (a)),

## 4.2 Ridge RMSE
Compute the out-of-sample RMSE for Ridge. Compare to both the simple OLS (16 vars) and the kitchen-sink OLS (~141 vars). What happened?
```{r}
# Compute test RMSE
pred_test_ridge <- predict(cv_ridge, s = lambda_ridge, newx = X_test)
rmse_test_ridge <- sqrt(mean((y_test - pred_test_ridge)^2) / nrow(X_test))
rmse_test_ridge
```

The out-of sample RMSE for Ridge is `rmse_test_ridge`, which is slightly lower than the out-of sample RMSE's for the simple model (`rmse_test_simple`) and the kitchen-sink model (`rmse_test_full`). This shows how Ridge using cross-validation is able to tune the model to have less prediction error than the other two models.

## 4.3 Did Ridge set any coeﬀicients to zero?
Extract the Ridge coeﬀicients using `coef(cv_ridge, s = "lambda.min").` Did Ridge eliminate any variables? Why or why not?
```{r}
ridge_coefs <- coef(cv_ridge, s = "lambda.min")
ridge_coefs
```

From the above, Ridge seemed to eliminate the interaction terms that were between racial categories, which makes sense since an individual can only be in 1 racial category, so these predictors are independent of each other and the interactions between these independent terms are then excluded. Similarly, Ridge excludes the interaction term of `age_under_25:age_25_45`, which again is because these categories are mutually exclusive.

\newpage

# 5 LASSO Regression: Shrink and Select (15 points)
LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty. Unlike Ridge, LASSO can shrink coeﬀicients all the way to exactly zero, effectively performing variable selection.

## 5.1 Fit LASSO with cross-validation
Use `cv.glmnet()` with alpha = 1 on the expanded training data. Remember to set your seed first for reproducibility.

```{r}
set.seed(51)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
```

(a) What is the optimal $\lambda$?
```{r}
lambda_lasso <- cv_lasso$lambda.min
lambda_lasso
```
The optimal $\lambda$ is `lambda_lasso`

(b) Plot the cross-validation curve.
```{r}
plot(cv_lasso)
```
This plot is showing the Mean-Squared Error values that are calculated from the LASSO model for different values of $\lambda$ on the $-log(\lambda)$ scale. The numbers at the top of the plot are the number of non-zero coefficients included in the model as the value of $-log(\lambda)$ increases, and it makes sense that this number increases as $-log(\lambda)$ increases since LASSO is very penalizing by setting coefficients exactly to 0. The two vertical dashed lines on the graph indicate the region of $-log(\lambda)$ values that provide the best model performance. From this plot, it appears that the optimal $-log(\lambda)$ is between 4.5 and 5 (meaning that $\lambda \approx e^{-4.75} \approx 0.008652$, which is close to the optimal $\lambda$ value that we found in (a)), when the model has ~27 coefficients.

## 5.2 LASSO variable selection
Extract the LASSO coeﬀicients at lambda.min.
(a) How many of the ~141 variables survived (non-zero coeﬀicient)? How many were eliminated?
```{r}
coef(cv_lasso, s = "lambda.min")
```
From the above, LASSO only kept 27 predictors and eliminated the rest of the 114 predictors, with a tendency to keep predictors and interactions between `female`, `age`, `age_under_25`, `priors_count`, `felony`, and categories having to do with juvy.  

(b) List a few of the surviving interaction terms. Do they make substantive sense?
A few of the surviving interaction terms are `juv_fel_count:age_under_25`, `juv_other_count:age_under_25`,  `has_juv_felony:age_under_25`, which logistically makes sense since these terms are saying that for individuals under 25, the effect for having a juvy felony or their juvy felony count increasing is greater than if they were in the 25-45 age category. Other categories show interactions between racial groups and having certain charges like `hispanic:juv_fel_count` and `black:juv_fel_count`, both of which have a positive effect and suggest possible racial bias in the way charges are distributed among different racial groups.

(c) Baker (2025) argues that LASSO reduces “subjective discretion” in model building. How does what you see here illustrate that argument?
LASSO appears to reduce “subjective discretion” by only keeping predictors who's inclusion is significant, whereas manually choosing which predictors to keep might reflect some bias on the researcher's part by either not including predictors that might actually have an effect (omitted variable bias), or, including variables that are not that significant but improve other metrics of the model suhc as $R^2$. As such, LASSO algorithmically and optimally chooses which predictors to keep rather than a researcher using solely their theoretical framework to decide.

## 5.3 LASSO RMSE
Compute the out-of-sample RMSE for LASSO. How does it compare to OLS (both versions) and Ridge?
```{r}
# Compute test RMSE
pred_test_lasso <- predict(cv_lasso, s = lambda_lasso, newx = X_test)
rmse_test_lasso <- sqrt(mean((y_test - pred_test_lasso)^2) / nrow(X_test))
rmse_test_lasso
```

The out-of sample RMSE for LASSO is `rmse_test_lasso`, making it the smallest one in comparison to the  out-of sample RMSE's for the simple model (`rmse_test_simple`), the kitchen-sink model (`rmse_test_full`), and even the Ridge model (`rmse_test_ridge`). Although overall, the LASSO and Ridge RMSE's are very comaprable to each other, the same way the simple and kitchen-sink model's are comparable to each other, and shows that Ridge and LASSO using cross-validation yields slightly better-tune models for our purposes. 

\newpage

# 6 Model Comparison and Interpretation (30 points)

## 6.1 Elastic Net
Fit an Elastic Net with `alpha = 0.5` on the expanded training data. Remember to set your seed first. Compute its out-of-sample RMSE and count how many variables it retains.

```{r}
# Fit Elastic Net
set.seed(51)
cv_enet <- cv.glmnet(X_train, y_train, alpha = 0.5, nfolds = 10)

# Compute test RMSE
pred_test_enet <- predict(cv_enet, s = lambda_ridge, newx = X_test)
rmse_test_enet <- sqrt(mean((y_test - pred_test_enet)^2) / nrow(X_test))
rmse_test_enet
```

## 6.2 The big comparison table
Fill in this table with your results:

(a) Which model has the lowest out-of-sample RMSE?
The LASSO model, with an out-of-sample RMSE of `rmse_test_lasso`

(b) Which model has the best in-sample fit? Is that the same model?
The kitchen-sink model , with an in-sample RMSE of rmse_train_full`

(c) In 2–3 sentences, explain what this table teaches about the bias-variance tradeoff.
BLANK

## 6.3 Baker article connection
In Baker (2025), two experts in the Halliburton securities case selected different peer firms and reached different conclusions about whether a stock drop was statistically significant.

(a) How did LASSO help narrow the disagreement between dueling experts?
BLANK

(b) Based on your experience with LASSO in this problem set, what aspect of the traditional OLS approach gives the analyst the most room for manipulation?
BLANK

## 6.4 Prediction vs. explanation
We used these models to predict `recidivism.` Suppose instead you wanted to know whether `priors_count` causes higher recidivism.

(a) Would the LASSO coeﬀicient on `priors_count` answer this causal question? Why or why not?
BLANK

(b) What is the difference between using regression for prediction versus using it for causal inference?
BLANK

## 6.5 Algorithmic fairness
Algorithms like COMPAS are used in real courtrooms to inform bail and sentencing decisions.

(a) Using your simple OLS model (16 vars), compute the mean predicted recidivism probability separately for Black defendants and white defendants in the test set. Is there a difference?
```{r}

```


