---
title: "GOV51 PSET3"
author: "\\small https://github.com/Steffny2004/gov51-ps3"
subtitle: "\\large Steffny Meegoda"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
	echo = TRUE,
	eval = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# 1 Setup and Data Exploration (10 points)

## 1.1 Load packages and data
Load the tidyverse, glmnet, and knitr packages. Then load the COMPAS dataset from
data/compas_clean.csv.

```{r}
library(tidyverse)
library(glmnet)
library(knitr)

compas <- read_csv("data/compas_clean.csv")
```

## 1.2 How many observations and variables are in the dataset? What is the unit of observation?


## 1.3 Examine the outcome variable
The variable recidivism equals 1 if the person was rearrested within two years and 0 otherwise.
(a) What is the overall recidivism rate in the data?
(b) Is this outcome continuous or binary? Why does this matter for how we interpret predictions from a linear model?

## 1.4 Summary statistics
Create a summary statistics table showing the mean and standard deviation of all variables in the dataset. Briefly note which variables have the most variation and which have the least.

## 1.5 Explore key predictos
Create a bar chart showing the recidivism rate by number of prior offenses (`priors_count`). Group `priors_count` into categories: 0, 1–2, 3–5, 6–10, and 11+. What pattern do you see? 

\newpage

# 2 OLS Baseline (15 points)

## 2.1 Train/test split
Set your seed to 51 for reproducibility. Split the data into a training set (80%) and a test set (20%).

```{r}
set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]
```

How many observations are in each set?

## 2.2 Fit OLS with main effects only
Fit an OLS regression of recidivism on all other variables in the training data using `lm(recidivism ~ ., data = train)`. This gives you a model with 16 predictors.
(a) Which predictors are statistically significant at the 5% level?
(b) Interpret the coeﬀicient on priors_count in one sentence.

## 2.3 In-sample vs. out-of-sample RMSE
Root Mean Squared Error (RMSE) measures average prediction error:
$$RMSE$$

(a) Compute the in-sample RMSE (predictions on the training data).
(b) Compute the out-of-sample RMSE (predictions on the test data).
(c) How large is the gap between in-sample and out-of-sample RMSE? Is OLS overfitting with 16 predictors?

## 2.4 Cross-validation by hand
A single train/test split depends on which observations happened to land in each group. Cross- validation averages over multiple splits to give a more stable estimate of out-of-sample performance.

Implement 5-fold cross-validation for the simple OLS model:

1. Randomly assign each training observation to one of 5 folds.
2. For each fold 𝑘 = 1, … , 5: fit OLS on the other 4 folds, predict fold 𝑘, compute RMSE.
3. Average the 5 RMSE values.

```{r}
# Assign folds
set.seed(51)
folds <- sample(rep(1:5, length.out = nrow(train)))

rmse_cv <- numeric(5)
for (k in 1:5) {
  train_k <- train[folds != k, ]
  test_k <- train[folds == k, ]
  model_k <- lm(recidivism ~
  ., data = train_k)
  pred_k <- predict(model_k, newdata = test_k)
  rmse_cv[k] <- sqrt(mean((test_k$recidivism - pred_k)^2))
}

mean(rmse_cv)
```

(a) What is the 5-fold CV estimate of RMSE? How does it compare to your single-split out-of-sample RMSE?
(b) In your own words, why is cross-validation more reliable than a single train/test split?
(c) When we use cv.glmnet() later, it does this same process internally to choose the penalty parameter $\lambda$. Why is cross-validation the right tool for choosing $\lambda$?

\newpage

# 3 The Kitchen Sink: What Happens When You Add Too Many Predictors? (15 points)
The simple OLS model uses 16 main effects. But maybe interactions between variables would help. What if being young AND having prior offenses is different from the sum of each effect alone? Let’s find out what happens when we throw in *everything.*

## 3.1 Build the expanded model
The formula below creates all pairwise interactions plus squared terms for the continuous variables. This expands our feature set from 16 to about 141 predictors.

```{r}
# Full interaction formula
f_full <- recidivism ~ (.)^2 +
  I(age^2) + I(priors_count^2) +
  I(juv_fel_count^2) + I(juv_misd_count^2) +
  I(juv_other_count^2)

# Create design matrices for glmnet (we'll need these later)
X_train <- model.matrix(f_full, data = train)[, -1]
X_test <- model.matrix(f_full, data = test)[, -1]
y_train <- train$recidivism
y_test <- test$recidivism

# How many features now?
ncol(X_train)
```

(a) How many predictors does this expanded model have?
(b) What is the ratio of training observations to predictors? (Compare to the simple model.)

## 3.2 Fit OLS with the kitchen sink
Fit OLS on the expanded feature set: ols_full <- lm(f_full, data = train).

(a) Compute the in-sample RMSE for this model.
(b) Compute the out-of-sample RMSE.
(c) How does the in-sample/out-of-sample gap compare to the simple 19-predictor OLS? What happened and why?

## 3.3 The overfitting lesson
In 2–3 sentences, explain why adding 180 interaction terms made the model’s **in-sample** fit better but its **out-of-sample** predictions worse. What is this phenomenon called?

\newpage 

# 4 Ridge Regression: Shrink Everything (15 points)

Ridge regression adds an L2 penalty that shrinks all coeﬀicients toward zero. Larger $\lambda$ means
more shrinkage. The key idea: Ridge trades a small amount of bias for a large reduction in
variance.

## 4.1 Fit Ridge with cross-validation
Use `cv.glmnet()` with `alpha = 0` to fit Ridge regression on the expanded training data (the X_train matrix with ~141 features). Use 10-fold cross-validation (the default) to select the optimal $\lambda$.

```{r}
set.seed(51) # for reproducible CV folds
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
```

(a) What is the optimal$\lambda$ selected by cross-validation? (Use `lambda.min`)
(b) Plot the cross-validation curve using `plot(cv_ridge)`. What does this plot show?

## 4.2 Ridge RMSE
Compute the out-of-sample RMSE for Ridge. Compare to both the simple OLS (16 vars) and the kitchen-sink OLS (~141 vars). What happened?

## 4.3 Did Ridge set any coeﬀicients to zero?
Extract the Ridge coeﬀicients using `coef(cv_ridge, s = "lambda.min").` Did Ridge eliminate any variables? Why or why not?

\newpage

# 5 LASSO Regression: Shrink and Select (15 points)
LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty. Unlike Ridge, LASSO can shrink coeﬀicients all the way to exactly zero, effectively performing variable selection.

## 5.1 Fit LASSO with cross-validation
Use `cv.glmnet()` with alpha = 1 on the expanded training data. Remember to set your seed first for reproducibility.

```{r}
set.seed(51)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
```

(a) What is the optimal $\lambda$?
(b) Plot the cross-validation curve.

## 5.2 LASSO variable selection
Extract the LASSO coeﬀicients at lambda.min.
(a) How many of the ~141 variables survived (non-zero coeﬀicient)? How many were eliminated?
(b) List a few of the surviving interaction terms. Do they make substantive sense?
(c) Baker (2025) argues that LASSO reduces “subjective discretion” in model building. How does what you see here illustrate that argument?

## 5.3 LASSO RMSE
Compute the out-of-sample RMSE for LASSO. How does it compare to OLS (both versions) and Ridge?

\newpage

# 6 Model Comparison and Interpretation (30 points)

## 6.1 Elastic Net
Fit an Elastic Net with `alpha = 0.5` on the expanded training data. Remember to set your seed first. Compute its out-of-sample RMSE and count how many variables it retains.

```{r}
set.seed(51)
cv_enet <- cv.glmnet(X_train, y_train, alpha = 0.5, nfolds = 10)
```

## 6.2 The big comparison table
Fill in this table with your results:

(a) Which model has the lowest out-of-sample RMSE?
(b) Which model has the best in-sample fit? Is that the same model?
(c) In 2–3 sentences, explain what this table teaches about the bias-variance tradeoff.

## 6.3 Baker article connection
In Baker (2025), two experts in the Halliburton securities case selected different peer firms and reached different conclusions about whether a stock drop was statistically significant.

(a) How did LASSO help narrow the disagreement between dueling experts?
(b) Based on your experience with LASSO in this problem set, what aspect of the traditional OLS approach gives the analyst the most room for manipulation?

## 6.4 Prediction vs. explanation
We used these models to predict recidivism. Suppose instead you wanted to know whether
priors_count causes higher recidivism.

(a) Would the LASSO coeﬀicient on priors_count answer this causal question? Why or why not?
(b) What is the difference between using regression for prediction versus using it for causal inference?

## 6.5 Algorithmic fairness
Algorithms like COMPAS are used in real courtrooms to inform bail and sentencing decisions.

(a) Using your simple OLS model (16 vars), compute the mean predicted recidivism prob- ability separately for Black defendants and white defendants in the test set. Is there a difference?

